\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-11-27 chapters/wcble.tex
Reply to DeLong test

\chapter{\Acs{erp} latency estimation and alignment}
\label{sec:wcble}
\epigraph{``The sands of time cannot be stopped''}{Brom in Eragon --
Christopher Paolini}
\noindent\emph{\Cref{sec:wcble/literature/contrib}, \cref{sec:wcble/methods} (including
\cref{fig:diagram-train,fig:diagram-eval}), and the third-to-last paragraph of
\cref{sec:wcble/conclusion} were published
as part of~\textcite{VanDenKerchove2024}.}

@@

\subsection{\acl{wcble}}
To improve performance over \ac{cble}, we propose a new algorithm inspired both by
\ac{cble} and the aforementioned Woody iteration scheme termed \ac{wcble}.
Instead of using \ac{cble} to estimate the features of a second-stage classifier
directly, \ac{cble} latency estimation is used as a step in a Woody iteration scheme.
While the Woody algorithm iteratively enhances the \ac{snr} of an \ac{erp} template to
cross-correlate with the data, \ac{wcble} iteratively re-estimates the parameters of
the first-stage classifier.
To improve convergence and perform well in a classification setting, \ac{wcble}
aligns both targets and non-targets to their corresponding estimated latencies.

The \ac{wcble} algorithm is presented in \cref{sec:wcble/app/wcble-alg}.
Its training phase is visualized in~\cref{fig:diagram-train}.
The initial training epochs $\{\mat{X}_n^{(1)}\}_{n=1}^N$ are set to $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N$.
At every iteration, classifier $\mathcal{C}$ is trained like in \ac{cble}:
\begin{equation}
  \theta^{(i)} =
  \mathrm{train}_\mathcal{C}(\{\mat{X}^{(i-1)}_n[:,s_1:s_2]\}_{n=1}^N,\mathbf{l}^\mathrm{train})
\end{equation}
Next, latency $s_{l_n}^{(i)}$ is determined for every epoch $\mat{X}^{(i)}$ corresponding
to its class label $l_n$ using \cref{eq:prob-latency}.
Finally, the training epochs $\mat{X}^{(i+1)}$ for the next iteration are determined by aligning
each original training epoch to the latency $s_{l_n}^{(i)}$ corresponding to its respective class
label.
\begin{equation}
  \mat{X}^{(i+1)}_n = \mathrm{align}(\mat{X}_n^\mathrm{train}, s_{l_n}^{(i)}) \quad \forall n=1,\ldots,N
\end{equation}
Aligning is performed by shifting and zero-padding the signal to the right if
the latency is negative relative to the time window onset, and to the left if
positive, by the difference between the latency and the window onset.
\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/wcble/figure1.pdf}
}{%
  \caption[Schematic representation of the \acs{wcble} training phase.]{%
    Schematic representation of the \acf{wcble} training phase.
    (1) The first-stage spatiotemporal binary classifier is trained on a set of
    epochs.
    (2) It is then applied to time-shifted copies of the epochs to obtain scores
    and class probabilities over time.
    (3) The medians of these probability distributions are assumed as the class
    latencies.
    (4) The epochs are aligned to their corresponding class latencies by
    shifting in time such that all latencies fall at the same moment.
    (5) The spatiotemporal classifier is then retrained on the aligned epochs
    for the next iteration.
    (6) After the iterative process halts, the scores and latencies obtained
    from the last iteration are used to train the second-stage
    classifier.
  }
  \label{fig:diagram-train}
}
The process halts after a fixed number of iterations or when the estimated set
of latencies has been encountered before, indicating it ended up in a loop.
In the end, the procedure should result in enhanced classifier parameters $\theta^*$,
closer to those when there would be no jitter between epochs.
Note that using the median approach for robust latency estimation results in a
smoother yet longer convergence process compared to the maximum/minimum approach.
We can then apply the classifier with enhanced parameters $\theta^*$ in a \ac{cble}
manner to unseen epochs as illustrated in \cref{fig:diagram-eval} to obtain a
vector of scores over time as in \cref{sec:wcble/methods/cble} and the estimated latencies as
in~\cref{sec:robust-latency}.
\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/wcble/figure2.pdf}
}{%
  \caption[Schematic representation of the \acs{wcble} test phase.]{%
    Schematic representation of the \acf{wcble}. (1) The
	  first-stage spatiotemporal binary classifier obtained from the training phase is
	  applied to time-shifted copies of the epochs to obtain scores and class probabilities
	  over time. (2) The medians of these probability distributions are assumed
	  as the new class latencies. (3) The scores and class latencies are input to the
    trained second-stage classifier, which predicts the label of the epochs.
  }
  \label{fig:diagram-eval}
}
\section{Results}
\label{sec:wcble/results}
Since it is hard to obtain ground truth measures of \ac{erp} latencies, we
first evaluate our approach with synthetic data.
In \cref{sec:covert-align}, the proposed algorithm will be applied to real
\ac{eeg} data.

\subsection{Synthetic data generation}
To verify our approach, we simulated synthetic 16 channel \ac{eeg} data with standard
10-20 electrode positions using the \texttt{simulate\_evoked} method of the
MNE (version 1.8)~\cite{Gramfort2013} software package.
\Acp{erp} were generated by projecting a sine wave pulse source time course in a dipole in the
left hemisphere to the scalp electrodes using a boundary element method forward
model~\cite{Mosher1999} and MNE's \texttt{fsaverage} source space and anatomy.
Pink temporal noise is generated by passing Gaussian noise through an infinite
impulse response filter with transfer function
$\frac{B(z)}{A(z)} = \frac{1}{1 -1z^{-1}+0.15z^{-2}}$.
Pink spatial noise is also added, using a noise covariance matrix constructed from the
cosine distances between electrode positions and scaling it by the inverse of
the resulting spatial covariance.
We refer to \textcite{Gramfort2014} for implementation details.

@@

>>>>

\chapter{\Acs{erp} latency estimation and alignment}
\label{sec:wcble}
\epigraph{``The sands of time cannot be stopped.''}{Brom in \emph{Eragon --
Christopher Paolini}}
\noindent\emph{\Cref{sec:wcble/literature/contrib},
\crefrange{sec:wcble/methods/cble}{sec:wcble/methods/wcble} (including
\cref{fig:diagram-train,fig:diagram-eval}), and the third-to-last paragraph of
\cref{sec:wcble/conclusion} were published as part of~\textcite{VanDenKerchove2024}.}

@@

\subsection{\acl{wcble}}
\label{sec:wcble/methods/wcble}
To improve performance over \ac{cble}, we propose a new algorithm inspired both by
\ac{cble} and the aforementioned Woody iteration scheme termed \ac{wcble}.
Instead of using \ac{cble} to estimate the features of a second-stage classifier
directly, \ac{cble} latency estimation is used as a step in a Woody iteration scheme.
While the Woody algorithm iteratively enhances the \ac{snr} of an \ac{erp} template to
cross-correlate with the data, \ac{wcble} iteratively re-estimates the parameters of
the first-stage classifier.
To improve convergence and perform well in a classification setting, \ac{wcble}
aligns both targets and non-targets to their corresponding estimated latencies.

The \ac{wcble} algorithm is presented in \cref{sec:wcble/app/wcble-alg}.
Its training phase is visualized in~\cref{fig:diagram-train}.
The initial training epochs $\{\mat{X}_n^{(1)}\}_{n=1}^N$ are set to $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N$.
At every iteration, classifier $\mathcal{C}$ is trained like in \ac{cble}:
\begin{equation}
  \theta^{(i)} =
  \mathrm{train}_\mathcal{C}(\{\mat{X}^{(i-1)}_n[:,s_1:s_2]\}_{n=1}^N,\mathbf{l}^\mathrm{train})
\end{equation}
Next, latency $s_{l_n}^{(i)}$ is determined for every epoch $\mat{X}^{(i)}$ corresponding
to its class label $l_n$ using \cref{eq:prob-latency}.
Finally, the training epochs $\mat{X}^{(i+1)}$ for the next iteration are determined by aligning
each original training epoch to the latency $s_{l_n}^{(i)}$ corresponding to its respective class
label.
\begin{equation}
  \mat{X}^{(i+1)}_n = \mathrm{align}(\mat{X}_n^\mathrm{train}, s_{l_n}^{(i)}) \quad \forall n=1,\ldots,N
\end{equation}
Aligning is performed by shifting and zero-padding the signal to the right if
the latency is negative relative to the time window onset, and to the left if
positive, by the difference between the latency and the window onset.
\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/wcble/figure1.pdf}
}{%
  \caption[Schematic representation of the \acs{wcble} training phase.]{%
    Schematic representation of the \acf{wcble} training phase.
    (1) The first-stage spatiotemporal binary classifier is trained on a set of
    epochs.
    (2) It is then applied to time-shifted copies of the epochs to obtain scores
    and class probabilities over time.
    (3) The medians of these probability distributions are assumed as the class
    latencies.
    (4) The epochs are aligned to their corresponding class latencies by
    shifting in time such that all latencies fall at the same moment.
    (5) The spatiotemporal classifier is then retrained on the aligned epochs
    for the next iteration.
    (6) After the iterative process halts, the scores and latencies obtained
    from the last iteration are used to train the second-stage
    classifier.
  }
  \label{fig:diagram-train}
}
The process halts after a fixed number of iterations or when the estimated set
of latencies has been encountered before, indicating it ended up in a loop.
In the end, the procedure should result in enhanced classifier parameters $\theta^*$,
closer to those when there would be no jitter between epochs.
Note that using the median approach for robust latency estimation results in a
smoother yet longer convergence process compared to the maximum/minimum approach.
We can then apply the classifier with enhanced parameters $\theta^*$ in a \ac{cble}
manner to unseen epochs as illustrated in \cref{fig:diagram-eval} to obtain a
vector of scores over time as in \cref{sec:wcble/methods/cble} and the estimated latencies as
in~\cref{sec:robust-latency}.
\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/wcble/figure2.pdf}
}{%
  \caption[Schematic representation of the \acs{wcble} test phase.]{%
    Schematic representation of the \acf{wcble}. (1) The
	  first-stage spatiotemporal binary classifier obtained from the training phase is
	  applied to time-shifted copies of the epochs to obtain scores and class probabilities
	  over time. (2) The medians of these probability distributions are assumed
	  as the new class latencies. (3) The scores and class latencies are input to the
    trained second-stage classifier, which predicts the label of the epochs.
  }
  \label{fig:diagram-eval}
}

\subsection{Synthetic data generation}
\label{sec:wcble/methods/simulation}
Since it is hard to obtain ground truth measures of \ac{erp} latencies, we
first evaluate our approach with synthetic data.
In \cref{sec:covert-align}, the proposed algorithm will be applied to real
\ac{eeg} data.

To verify our approach, we simulated synthetic 16 channel \ac{eeg} data with standard
10-20 electrode positions using the \texttt{simulate\_evoked} method of the
MNE (version 1.8)~\cite{Gramfort2013} software package.
\Acp{erp} were generated by projecting a sine wave pulse source time course in a dipole in the
left hemisphere to the scalp electrodes using a boundary element method forward
model~\cite{Mosher1999} and MNE's \texttt{fsaverage} source space and anatomy.
Pink temporal noise is generated by passing Gaussian noise through an infinite
impulse response filter with transfer function
$\frac{B(z)}{A(z)} = \frac{1}{1 -1z^{-1}+0.15z^{-2}}$.
Pink spatial noise is also added, using a noise covariance matrix constructed from the
cosine distances between electrode positions and scaling it by the inverse of
the resulting spatial covariance.
We refer to \textcite{Gramfort2014} for implementation details.

@@

\section{Results}
\label{sec:wcble/results}

### 2024-11-27 figures/bci/paradigms.tikz.tex
Resturcture chap 1 and 2
Rewrite sota
funding
Indicated comments Reinhold and Fabien

\begin{tikzpicture}[scale=\textwidth/2cm]
    \pgfmathsetmacro{\margin}{.1}
    \pgfmathsetmacro{\center}{.5+.5*\margin}
    \pgfmathsetmacro{\textspacing}{.15}

    % Draw axes
    \draw[darkgray, very thick, <->] ({-(1+\margin)}, 0) -- ({1+\margin},0);
    \draw[darkgray, very thick, <->] (0,{-(1+\margin)}) -- (0,{1+\margin});

@@

    %% Add text
    \node[color=accent1, align=left,font=\bfseries, anchor=north west] at  (-1,1){active};
    \node[color=accent2, align=left,font=\bfseries, anchor=north east] at  (1,1){reactive};
    \node[color=accent3, align=left,font=\bfseries, anchor=south west] at  (-1,-1){passive};

@@

    %% Text in rectangles
    \node[align=center] at ({-\center}, {\center+\textspacing}) {motor\\imagery};
    \node[align=center] at ({-\center}, {\center-\textspacing}) {imagined\\speech};

    \node[align=center] at ({\center+\textspacing}, {\center+\textspacing}) {P300};
    \node[align=center] at ({\center+\textspacing}, {\center-\textspacing}) {SSVEP};
    \node[align=center] at ({\center-\textspacing}, {\center+\textspacing}) {cVEP};
    \node[align=center] at ({\center-\textspacing}, {\center-\textspacing}) {mVEP};

    %% More text
    \node[align=center] at (\center, {-\center}) {error\\potentials};
    \node[align=center] at ({-\center}, {-\center}) {attention\\monitoring};
    \node[align=center] at (-.25, -.8) {emotion\\monitoring};
    \node[align=center] at (.25, -.8) {media induced\\emotion recognition};
\end{tikzpicture}

>>>>

\sffamily
\begin{tikzpicture}[
    scale=\textwidth/3cm,
  ]
  \useasboundingbox (-1,-1.25) rectangle (1,1);
  \pgfmathsetmacro{\margin}{.1}
  \pgfmathsetmacro{\center}{.5+.5*\margin}
  \pgfmathsetmacro{\textspacing}{.15}
  %\begin{scope}[shift={(1,1.2)}]
    % Draw axes
    \draw[darkgray, very thick, <->] ({-(1+\margin)}, 0) -- ({1+\margin},0);
    \draw[darkgray, very thick, <->] (0,{-(1+\margin)}) -- (0,{1+\margin});

@@

    %% Add text
    \node[color=accent1, align=left,font=\bfseries, anchor=north west, inner sep=2pt] at  (-1,1){active};
    \node[color=accent2, align=left,font=\bfseries, anchor=north east, inner sep=2pt] at  (1,1){reactive};
    \node[color=accent3, align=left,font=\bfseries, anchor=south west, inner sep=2pt] at  (-1,-1){passive};

@@

    %% Text in rectangles
    \node[align=center,font=\footnotesize, color=muteblack] at ({-\center}, {\center+1.5*\textspacing}) {motor\\imagery};
    \node[align=center,font=\footnotesize, color=muteblack] at ({-\center}, {\center}) {imagined\\speech};
    \node[align=center,font=\footnotesize, color=muteblack] at ({-\center}, {\center-1.5*\textspacing}) {neurofeedback};

    \node[align=center,font=\footnotesize, color=muteblack] at ({\center+\textspacing}, {\center+\textspacing}) {P300};
    \node[align=center,font=\footnotesize, color=muteblack] at ({\center+\textspacing}, {\center-\textspacing}) {SSVEP};
    \node[align=center,font=\footnotesize, color=muteblack] at ({\center-\textspacing}, {\center+\textspacing}) {cVEP};
    \node[align=center,font=\footnotesize, color=muteblack] at ({\center-\textspacing}, {\center-\textspacing}) {mVEP};

    %% More text
    \node[align=center,font=\footnotesize, color=muteblack] at ({\center}, {-\center+\textspacing}) {error\\potentials};
    \node[align=center,font=\footnotesize, color=muteblack] at ({-\center}, {-\center+\textspacing}) {attention and \\ workload detection};
    \node[align=center,font=\footnotesize, color=muteblack] at ({-\center}, {-\center-\textspacing}) {clinical neuroimaging \\ and monitoring};
    \node[align=center,font=\footnotesize, color=muteblack] at ({+\center}, {-\center-\textspacing})  {emotion \\ detection};
  %\end{scope}
\end{tikzpicture}
